---
title: "lab 3: COVID-19"
subtitle: 'Ecosystem Science and Sustainability 330'
author:
  - name: Megan Hoover
    email: megan.hoover21@gmail.com
format: html
execute: 
  echo: true
output:
  html
    self contained:true
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

## Q2: Lets pretend it in January 1st, 2022. You are a data scientist for the state of Colorado Department of Public Health.

You’ve been tasked with giving a report to Governor Polis each morning about the most current COVID-19 conditions at the county level.

As it stands, the Colorado Department of Public Health maintains a watch list of counties that are being monitored for worsening corona virus trends. There are six criteria used to place counties on the watch list:

1.Doing fewer than 150 tests per 100,000 residents daily (over a 7-day average)
2.More than 100 new cases per 100,000 residents over the past 14 days…
3.25 new cases per 100,000 residents and an 8% test positivity rate
4.10% or greater increase in COVID-19 hospitalized patients over the past 3 days
5.Fewer than 20% of ICU beds available
6.Fewer than 25% ventilators available 
Of these 6 conditions, you are in charge of monitoring condition number 2.

To do this job well, you should set up a reproducible framework to communicate the following in a way that can be updated every time new data is released (daily):

1. cumulative cases in the 5 worst counties
2. total NEW cases in the 5 worst counties
3. A list of safe counties
4. A text report describing the total new cases, total cumulative cases, and number of safe counties.
You should build this analysis in such a way that running it will extract the most current data straight from the NY-Times URL and the state name and date are parameters that can be changed allowing this report to be run for other states/dates.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r} 
#Download libraries,read in COVID-19 Data, and prepare data table for calculations
library(tidyverse) #data wrangling and visualization) 

library(flextable) #make nice tables)

library(zoo) #(rolling averages

url = 'https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv'
covid_raw<-read_csv(url)

#create my.date and my.state object. Set my date to "2022-02-01" and my state to "colorado"
my.date <- "2022-02-01" 
  class(my.date)
date<-my.date %>%
  as.Date()
class(date)
my.state <- "Colorado"

#Start by making a subset that limits (filter) the data to Colorado and add a new column (mutate) with the daily new cases using diff/lag by county (group_by). Do the same for new deaths as well.

#(Hint: you will need some combination of filter, group_by, arrange, mutate, diff/lag, and ungroup)

covid1<-covid_raw %>%
  filter(state=="Colorado") %>%
  drop_na() %>%
  group_by(county) %>%
  mutate(daily_new_cases= cases-lag(cases),
         daily_new_deaths= deaths- lag(deaths)) %>%
  arrange(county,date) %>%
  ungroup()
```

